---
  output: github_document
---

  <!-- README.md is generated from README.Rmd. Please edit that file -->

  ```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```
# groupedstats: Grouped statistical analysis in a tidy way

[![packageversion](https://img.shields.io/badge/Package%20version-0.0.1.9000-orange.svg?style=flat-square)](commits/master)
[![Travis Build Status](https://travis-ci.org/IndrajeetPatil/groupedstats.svg?branch=master)](https://travis-ci.org/IndrajeetPatil/groupedstats) 
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/IndrajeetPatil/groupedstats?branch=master&svg=true)](https://ci.appveyor.com/project/IndrajeetPatil/groupedstats) 
[![Licence](https://img.shields.io/badge/licence-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-red.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.3.0-6666ff.svg)](https://cran.r-project.org/)
<!-- [![Coverage Status](https://img.shields.io/codecov/c/github/IndrajeetPatil/groupedstats/master.svg)](https://codecov.io/github/IndrajeetPatil/groupedstats?branch=master)
[![Dependency Status](http://img.shields.io/gemnasium/IndrajeetPatil/groupedstats.svg)](https://gemnasium.com/IndrajeetPatil/groupedstats) -->

## Overview

`groupedstats` package provides a collection of functions to run statistical
operations on multiple variables across multiple grouping variables in a
dataframe.

## Installation

You can get the development version from GitHub. If you are in hurry and want to reduce the time of installation, prefer-

```{r installation1, eval = FALSE}
# install.packages("devtools")                                # needed package to download from GitHub repo
devtools::install_github(repo = "IndrajeetPatil/groupedstats", # package path on GitHub
                         quick = TRUE)                        # skips docs, demos, and vignettes
```

If time is not a constraint-
```{r installation2, eval = FALSE}
devtools::install_github(repo = "IndrajeetPatil/groupedstats", # package path on GitHub
                         dependencies = TRUE,                 # installs packages which groupedstats depends on
                         upgrade_dependencies = TRUE          # updates any out of date dependencies
)
```

## Help

Documentation for any function can be accessed with the standard `help` command-
```{r help, eval = FALSE}
?grouped_lm
?grouped_robustlm
?grouped_proptest
?grouped_summary
```

## Usage

* `grouped_summary`

Getting summary for multiple variables across multiple grouping variables.

```{r grouped_summary}
library(datasets)

groupedstats::grouped_summary(data = mtcars,
                              grouping.vars = c(am, cyl),
                              measures = c(disp, wt, mpg))
```

* `grouped_lm`

This function can be used to run linear regression between different pairs of variables across multiple levels of grouping variables. For example, if we want to assess the linear relationship between `Sepal.Length` and `Sepal.Width`, **and** `Sepal.Length` and `Sepal.Width` (notice the order in which variables are being entered) for all levels of `Species`:    

```{r grouped_lm1}
library(datasets)

groupedstats::grouped_lm(data = iris,
                         crit.vars = c(Sepal.Length, Petal.Length),
                         pred.vars = c(Sepal.Width, Petal.Width),
                         grouping.vars = Species)
```

This can be done with multiple grouping variables. For example, in the following
example, we use the `gapminder` dataset to regress life expectency and
population on GDP per capita for each continent and for each country.

```{r grouped_lm2, warning = FALSE, message = FALSE}
library(gapminder)
library(dplyr)

groupedstats::grouped_lm(data = gapminder::gapminder,
                         crit.vars = c(lifeExp, pop),
                         pred.vars = c(gdpPercap, gdpPercap),
                         grouping.vars = c(continent, country)) %>%
  dplyr::arrange(.data = ., continent, country)
```

* `grouped_robustlm`

There is also robust variant of linear regression (as implemented in `robust::lmRob`)-

```{r grouped_robustlm, warning = FALSE, message = FALSE}
library(gapminder)
library(dplyr)

groupedstats::grouped_robustlm(data = gapminder::gapminder,
                         crit.vars = c(lifeExp, pop),
                         pred.vars = c(gdpPercap, gdpPercap),
                         grouping.vars = c(continent, country)) %>%
  dplyr::arrange(.data = ., continent, country)
```

* `grouped_proptest`

This function helps carry out one-sample proportion tests with a unique variable for multiple grouping variables-

```{r grouped_proptest}
library(datasets)

groupedstats::grouped_proptest(data = mtcars,
                               grouping.vars = cyl,
                               measure = am)
```
