---
  output: github_document
---

  <!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, warning = FALSE, messsage = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  out.width = "100%",
  comment = "#>",
  fig.path = "man/figures/README-"
)
```
# groupedstats: Grouped statistical analysis in a tidy way

[![CRAN_Release_Badge](http://www.r-pkg.org/badges/version-ago/groupedstats)](https://CRAN.R-project.org/package=groupedstats)
[![CRAN Checks](https://cranchecks.info/badges/summary/groupedstats)](https://cran.r-project.org/web/checks/check_results_groupedstats.html)
[![packageversion](https://img.shields.io/badge/Package%20version-0.0.1-orange.svg?style=flat-square)](commits/master)
[![Daily downloads badge](https://cranlogs.r-pkg.org/badges/last-day/groupedstats?color=blue)](https://CRAN.R-project.org/package=groupedstats)
[![Weekly downloads badge](https://cranlogs.r-pkg.org/badges/last-week/groupedstats?color=blue)](https://CRAN.R-project.org/package=groupedstats)
[![Monthly downloads badge](https://cranlogs.r-pkg.org/badges/last-month/groupedstats?color=blue)](https://CRAN.R-project.org/package=groupedstats)
[![Total downloads badge](https://cranlogs.r-pkg.org/badges/grand-total/groupedstats?color=blue)](https://CRAN.R-project.org/package=groupedstats)
[![Travis Build Status](https://travis-ci.org/IndrajeetPatil/groupedstats.svg?branch=master)](https://travis-ci.org/IndrajeetPatil/groupedstats) 
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/IndrajeetPatil/groupedstats?branch=master&svg=true)](https://ci.appveyor.com/project/IndrajeetPatil/groupedstats) 
[![Licence](https://img.shields.io/badge/licence-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-red.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.3.0-6666ff.svg)](https://cran.r-project.org/)
[![Coverage Status](https://img.shields.io/codecov/c/github/IndrajeetPatil/groupedstats/master.svg)](https://codecov.io/github/IndrajeetPatil/groupedstats?branch=master)

## Overview

`groupedstats` package provides a collection of functions to run statistical
operations on multiple variables across multiple grouping variables in a
dataframe. This is a common situation, as illustrated by few example cases-

1.  If you have combined multiple studies in a single dataframe and want to run
    a common operation (e.g., linear regression) on **each study**. In this
    case, column corresponding to `study` will be the grouping variable.
2.  If you have multiple groups in your dataframe (e.g., clinical disorder
    groups and controls group) and you want to carry out the same operation for
    **each group** (e.g., planned t-test to check for differences in reaction
    time in condition 1 versus condition 2 for both groups). In this case,
    `group` will be the grouping variable.
3.  If you have multiple conditions in a given study (e.g., six types of videos
    participants saw) and want to run a common operation between different
    measures of interest for **each condition** (e.g., correlation between
    subjective rating of emotional intensity and reaction time).
4. Combination of all of the above.

This is first installment of the package that provides only the most basic
statistical operations. The next releases will expand on the existing
functionality. 

## Installation

To get the latest, stable CRAN release (0.0.1):
```{r installation1, eval = FALSE, warning = FALSE, messsage = FALSE}
utils::install.packages(pkgs = "groupedstats") 
```

You can get the development version from GitHub. If you are in hurry and want to
reduce the time of installation, prefer-

```{r installation2, eval = FALSE, warning = FALSE, messsage = FALSE}
# needed package to download from GitHub repo
utils::install.packages(pkgs = "devtools")                 

devtools::install_github(repo = "IndrajeetPatil/groupedstats",  # package path on GitHub
                         quick = TRUE)                          # skips docs, demos, and vignettes
```

If time is not a constraint-
```{r installation3, eval = FALSE, warning = FALSE, messsage = FALSE}
devtools::install_github(repo = "IndrajeetPatil/groupedstats", # package path on GitHub
                         dependencies = TRUE,                  # installs packages which groupedstats depends on
                         upgrade_dependencies = TRUE           # updates any out of date dependencies
)
```

## Help

Documentation for any function can be accessed with the standard `help` command-
```{r help, eval = FALSE, warning = FALSE, messsage = FALSE}
?grouped_lm
?grouped_lmer
?grouped_summary
?grouped_slr
?grouped_robustslr
?grouped_proptest
?grouped_ttest
?grouped_wilcox
```

## Usage

* `grouped_summary`

Getting summary for multiple variables across multiple grouping variables. This
function is a wrapper around `skimr::skim_to_wide()`. It is supposed to be a
handy summarizing tool if you have multiple grouping variables and multiple
variables for which summary statistics are to be computed-

```{r grouped_summary1, warning = FALSE, messsage = FALSE}
library(datasets)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_summary(data = datasets::iris,
                              grouping.vars = Species,
                              measures = Sepal.Length:Petal.Width,
                              measures.type = "numeric")
```

This function can be used to get summary of either numeric **or** factor
varibles, but **not** both. This is by design. If no `measures` are specified, the function will
compute summary for all variables of the specified type (`numeric` or `factor`).

If you want summary of variables of `factor` type-

```{r grouped_summary2, warning = FALSE, messsage = FALSE}
library(ggplot2)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_summary(data = ggplot2::diamonds,
                              grouping.vars = c(cut, clarity),
                              measures = color,
                              measures.type = "factor")
```

Note that there is a column corresponding to `top_counts` which is really useful
in case you, let's say, want to plot these counts. But this column is of
`character` type and in wide format. The solution is to use an additional
argument provided for this function:

```{r grouped_summary3, fig.width = 8, warning = FALSE, messsage = FALSE}
library(ggplot2)
library(magrittr)
library(ggstatsplot)

options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_summary(
  data = ggplot2::diamonds,
  grouping.vars = cut,                          # for simplicity, let's just use one grouping variable
  measures = color,
  measures.type = "factor",
  topcount.long = TRUE
) %>% 
  ggplot2::ggplot(
  data = .,                                     # placeholder for summary dataframe we just created 
  mapping = ggplot2::aes(
    x = forcats::fct_inorder(f = factor.level),
    y = count,
    fill = factor.level
  )
) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::labs(x = "color", y = "count") +
  ggplot2::facet_grid(facets = ~ cut) +         # for each level of the factor level
  ggstatsplot::theme_mprl() +
  ggplot2::theme(legend.position = "none")
```

This produces a long format table with two new columns `factor.level` and its
corresponding `count`, which can then be immediately fed into other pipelines,
e.g., preparing a plot of `mean` and `sd` values in `ggplot2`).

```{r grouped_summary4, warning = FALSE, messsage = FALSE}
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_summary(
  data = ggplot2::diamonds,
  grouping.vars = c(cut, clarity)
)
```

* `grouped_slr`

This function can be used to run **simple linear regression** (slr) between
different pairs of variables across multiple levels of grouping variable(s). For
example, we can use the `gapminder` dataset to study two relationships of
interest for **each country** across years:

1.  life expectency and GDP (per capita)
2.  population GDP (per capita) Thus, in this case we have two regression models
    and one grouping variable with 142 levels (countries)

```{r grouped_slr1, warning = FALSE, messsage = FALSE}
library(gapminder)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_slr(data = gapminder::gapminder,
                         dep.vars = c(lifeExp, pop),
                         indep.vars = c(gdpPercap, gdpPercap),
                         grouping.vars = country)
```

Notice the order in which the dependent and independent variables are entered;
there are two regression models being run here: `lifeExp ~ gdpPercap` and `pop ~
gdpPercap` If this order is incorrect, the result will also be incorrect. So it
is always a good idea to check the *formula* column to see if you have run the
correct linear models. Also, note that the estimates are already standardized,
i.e. estimates are standardized regression coefficients (betas, i.e.). 

The prior example was with just one grouping variable. This can be done with multiple
grouping variables as well. For example, with the `diamonds` dataset from
`ggplot2` library, let's assess the relation between carat *and* price of a
diamond for each type of clarity *and* cut-

```{r grouped_slr2, warning = FALSE, messsage = FALSE}
library(ggplot2)
library(dplyr)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_slr(data = ggplot2::diamonds,
                         dep.vars = price,
                         indep.vars = carat,
                         grouping.vars = c(cut, clarity)) %>%
  dplyr::arrange(.data = ., cut)
```

A more general version of this function (`grouped_lm`) will be implemented in
future that will utilize the formula interface of `stats::lm`.

* `grouped_robustslr`

There is also robust variant of **simple linear regression** (as implemented in `robust::lmRob`)-

```{r grouped_robustslr, warning = FALSE, messsage = FALSE}
library(gapminder)
library(dplyr)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_robustslr(data = gapminder::gapminder,
                         dep.vars = c(lifeExp, pop),
                         indep.vars = c(gdpPercap, gdpPercap),
                         grouping.vars = c(continent, country)) %>%
  dplyr::arrange(.data = ., continent, country)
```
A more general version of this function (`grouped_robustlm`) will be implemented in
future that will utilize the formula interface of `robust::lmRob`.

* `grouped_lm`

A more general version of simple linear regression is `stats::lm`, implemented
in `grouped_lm`:

```{r grouped_lm1, warning = FALSE, messsage = FALSE}
library(groupedstats)

groupedstats::grouped_lm(
  data = mtcars,
  grouping.vars = cyl,        # grouping variable (just one in this case)
  formula = mpg ~ am*wt,      # note that this function takes a formula
  output = "tidy"             # tidy dataframe containing results
)
```

The same function can also be used to get model summaries instead of tidy
dataframe containing results-

```{r grouped_lm2, warning = FALSE, messsage = FALSE}
library(groupedstats)

groupedstats::grouped_lm(
  data = ggplot2::diamonds,
  grouping.vars = c(cut, color),            # grouping variables
  formula = price ~ carat * clarity,        # formula
  output = "glance"                         # dataframe with model summaries
)
```

* `grouped_aov`

A related function to `stats::lm` is `stats::aov`, which fits an analysis of
variance model for each group. Contrast the output you get here with the
previous output for the same model from `grouped_lm` function. The estimate in
this case with be an effect size (either partial eta-squared or partial
omega-squared).

```{r grouped_aov1, warning = FALSE, messsage = FALSE}
library(groupedstats)

groupedstats::grouped_aov(
  data = mtcars,
  grouping.vars = cyl,        # grouping variable (just one in this case)
  formula = mpg ~ am*wt       # note that this function takes a formula
)
```

* `grouped_glm`

The option to run generalized linear model across different levels of the grouping variable-

```{r grouped_glm, warning = FALSE, messsage = FALSE}
groupedstats::grouped_glm(data = datasets::mtcars,
                          dep.vars = am,
                          indep.vars = wt,
                          grouping.vars = cyl,
                          family = "gaussian")
```

* `grouped_lmer`

Linear mixed effects analyses (`lme4::lmer`) for all combinations of grouping
variable levels can be carried out using `grouped_lmer`:

```{r grouped_lmer1, warning = FALSE, messsage = FALSE}
library(gapminder)

# getting tidy output of results
groupedstats::grouped_lmer(
  data = gapminder,
  formula = scale(lifeExp) ~ scale(gdpPercap) + (gdpPercap |
                                                   continent),
  grouping.vars = year,
  REML = FALSE,
  output = "tidy"
)
```

* `grouped_glmer`

A more generalized version of `lmer` is implemented in `glmer`, which can also
handle categorical/nominal data. For example, let's say we want to see if sex of
a person was predictive of whether they survived the Titanic tragedy

```{r grouped_glmer1, warning = FALSE, messsage = FALSE}
# having a look at the data
dplyr::glimpse(x = groupedstats::Titanic_full)

# running glmer model to get tidy output
groupedstats::grouped_glmer(
  formula = Survived ~ Age + (Age |
                                Class),
  data = groupedstats::Titanic_full,
  family = stats::binomial(link = "probit"),                  # choosing the appropriate GLM family
                          control = lme4::glmerControl(       # choosing appropriate control
                            optimizer = "Nelder_Mead",
                            boundary.tol = 1e-07,
                            calc.derivs = FALSE,
                            optCtrl = list(maxfun = 2e9)
                          ),
  grouping.vars = Sex,                                        # grouping variables (just one in this case)
  output = "tidy"
)

# getting glmer model summaries (let's use the default family and control values)
groupedstats::grouped_glmer(
  formula = Survived ~ Age + (Age |
                                Class),
  grouping.vars = Sex,      
  data = groupedstats::Titanic_full,
  output = "glance"
)
```

* `grouped_proptest`

This function helps carry out one-sample proportion tests with a unique variable
for multiple grouping variables-

```{r grouped_proptest, warning = FALSE, messsage = FALSE}
library(datasets)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_proptest(data = mtcars,
                               grouping.vars = cyl,
                               measure = am)
```

* `grouped_ttest`

This function can help you carry out t-tests, paired or independent, on multiple
variables across multiple groups. Deomostrating how to use this function is
going to first require getting the `iris` dataset into long format. Let's say we
want to investigate if `Sepal` part of the flower has greater measurement
(length or width) than `Petal` part of the flower for **each** *Iris* species.

```{r grouped_ttest1, warning = FALSE, messsage = FALSE}

# converting the iris dataset to long format
iris_long <- datasets::iris %>%
  dplyr::mutate(.data = ., id = dplyr::row_number(x = Species)) %>%
  tidyr::gather(
    data = .,
    key = "condition",
    value = "value",
    Sepal.Length:Petal.Width,
    convert = TRUE,
    factor_key = TRUE
  ) %>%
  tidyr::separate(
    col = "condition",
    into = c("part", "measure"),
    sep = "\\.",
    convert = TRUE
  ) %>%
  tibble::as_data_frame(x = .)

# check the long format iris dataset
iris_long

# checking if the Sepal part has different dimentions (value) than Petal part
# for each Species and for each type of measurement (Length and Width)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_ttest(
  data = iris_long,
  dep.vars = value,                    # dependent variable
  indep.vars = part,                   # independent variable
  grouping.vars = c(Species, measure), # for each Species and for each measurement
  paired = TRUE                        # paired t-test
)
```

* `grouped_wilcox`

This function is just a non-parametric variant of the `grouped_ttest`:

```{r grouped_wilcox, warning = FALSE, messsage = FALSE}
# checking if the Sepal part has different dimentions (value) than Petal part
# for each Species and for each type of measurement (Length and Width)
options(tibble.width = Inf)            # show me all columns

groupedstats::grouped_wilcox(
  data = iris_long,
  dep.vars = value,                    # dependent variable
  indep.vars = part,                   # independent variable
  grouping.vars = c(Species, measure), # for each Species and for each measurement
  paired = TRUE                        # paired Wilcoxon signed rank test with continuity correction
)
```

While we are at it, let's also check out examples for `t-test` and `Wilcox test`
in case of between-subjects designs.

We will use `diamonds` dataset from `ggplot2` and will see if the price and
depth of a diamond is different for two of our favorite colors (say `E` and `J`)
for each type of clarity.

```{r grouped_wilcox2, warning = FALSE, messsage = FALSE}

# subset the dataframe with two colors of interest to us
diamonds_short <-
  dplyr::filter(.data = ggplot2::diamonds, color == "E" |
                  color == "J")

options(tibble.width = Inf, tibble.print_max = Inf)             # show me all rows and columns

# t-test
groupedstats::grouped_ttest(
  data = diamonds_short,
  dep.vars = c(carat, price, depth),             # note that there three dependent variables 
  indep.vars = color,                            # and just one independent variable 
  grouping.vars = clarity,                       # one grouping variable
  paired = FALSE,
  var.equal = FALSE
)

# wilcox test (aka Mann-Whitney U-test)
groupedstats::grouped_wilcox(
  data = diamonds_short,
  dep.vars = depth:price,                        # note that you can select variables in range with `:`
  indep.vars = color,                            # again, just one independent, multiple dependent variables case
  grouping.vars = clarity,                       # one grouping variable
  paired = FALSE
)
```

We can further focus just on two levels of clarity to further elucidate another
aspect of entering the arguments-

```{r grouped_wilcox3, warning = FALSE, messsage = FALSE}
# subset the dataframe even further to just select two levels of clarity
diamonds_short2 <-
  dplyr::filter(.data = diamonds_short, clarity == "SI2" |
                  clarity == "SI1")

# wilcox test (aka Mann-Whitney U-test)
groupedstats::grouped_wilcox(
 data = diamonds_short2,
  dep.vars = c(carat, price),                    # two dependent variables
  indep.vars = c(color, clarity),                # two independent variables
  grouping.vars = cut,                           # one grouping variable
  paired = FALSE
)

```

In these examples, two things are worth noting that generalize to **all**
functions in this package and stem from how [`tidy evaluation`](https://adv-r.hadley.nz/evaluation.html) works: <br />
 **1.** If just one independent variable is provided for multiple dependent
variables, it will be used as a common variable.  <br />
 **2.** If you want to use a selection of variables, you need not use `c()`.
 
* Suggestions

If you find any bugs or have any suggestions/remarks, please file an issue on
GitHub: <https://github.com/IndrajeetPatil/groupedstats/issues>

